---
title: "Kaessmann Data - YARN - TissueSpecificity - V7"
output:
  html_document:
      toc: TRUE
      toc_float: TRUE
      keep_md: FALSE
---

```{r setup, include=T, echo=T,error=F,warning=F,message=F}
#######
# Who: Siva
# When: Jun 2023
# What:
# - Split the pre- and post- natal samples separately and then perform tissue-specific normalization
# - Filtering of low expressed genes is also done based on the pre-post split set of samples.
# - Merge the Forebrain and Hindbrain samples
# - Remove testis and ovary
# - preferential expression (TiP) calculated using Eqn. 4 and 5 of publication https://doi.org/10.1016/j.celrep.2017.10.001 - Two diff formula used;
# - TiP4: 
# - ((median of all pre-natal brain) - (median of all prenatal non-brain)/(IQR of all prenatal samples))
#######

knitr::opts_chunk$set(echo = F, cache = F, warning = F)
options(stringsAsFactors = F)

library(yarn)
library(readr)
library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(reshape2)
library(Biobase)
library(kableExtra)
library(data.table)
library(stringr)
library(ggrepel)
library(multidplyr)

```

## Import raw read counts

We import the read counts (.gct files) for all samples obtained (processing info in prologue). We want to prepare the data in a format suited for making it as an expression set object. 

```{r import_raw, include=T, echo=T, error=F, warning=F, message=F}
load("./kaessmann_raw_expSetObj.RData")
```

## The YARN pipeline

### Filtering samples

As we agreed in our initial meetings to remove testis from all analyses, I remove them from subsequent processing.
We filter out genes from subsequent analysis which are low expressed (at least 1 read in half of the samples per tissue). 

```{r yarn_filter, include=T, echo=T, error=F, warning=F, message=F}
kaessmann_filteredouttestis = filterSamples(kaessmann_raw, ids = c("testis","ovary"), groups = "tissue")

# We want to add additional metadata, categorizing samples into pre-vs-postnatal
metadata_prevspost = as.data.frame(pData(kaessmann_filteredouttestis))
metadata_prevspost$natalgroup = "Postnatal"
metadata_prevspost[grepl("^[0-9]",metadata_prevspost$devstage),]$natalgroup = "Prenatal"

# We merge the fore and hind brain as brain
metadata_prevspost$tissue_original = metadata_prevspost$tissue
metadata_prevspost[grepl("^fore",metadata_prevspost$tissue),]$tissue = "brain"
metadata_prevspost[grepl("^hind",metadata_prevspost$tissue),]$tissue = "brain"

# reassign to the exp. set object
pData(kaessmann_filteredouttestis) = metadata_prevspost

# split the pre and post natal samples
kmann_prenatal = filterSamples(kaessmann_filteredouttestis, ids = "Postnatal", groups = "natalgroup")
kmann_postnatal = filterSamples(kaessmann_filteredouttestis, ids = "Prenatal", groups = "natalgroup")

# Perform the expression filter also in tissue level
kmann_prenatal_filtered = filterLowGenes(kmann_prenatal, "tissue")
kmann_postnatal_filtered = filterLowGenes(kmann_postnatal, "tissue")

```

The number of genes in our data set before and after filtering are `r dim(kaessmann_raw)[1]` and `r dim(kmann_prenatal_filtered)[1]`, `r dim(kmann_postnatal_filtered)[1]` in prenatal and postnatal respectively.

### Tissue-Natal State aware normalization

```{r yarn_denplot_after_norm, include=T, echo=T, error=F, warning=F, message=F}

# Normalize
kmann_prenatal_filtered <- normalizeTissueAware(kmann_prenatal_filtered,"tissue")
kmann_postnatal_filtered <- normalizeTissueAware(kmann_postnatal_filtered,"tissue")

# get required columns and join them from assayData and phenoData of the ExpressionSet object
kmann_prenatal_filtered_normcount_meta_df <- assayData(kmann_prenatal_filtered)[["normalizedMatrix"]] %>% 
                    as.data.frame() %>%
                    rownames_to_column(., var = "gene_id") %>%
                    pivot_longer(cols = -c("gene_id"), names_to = "SAMPID", values_to = "norm_count") %>%
                    left_join(pData(kmann_prenatal_filtered) %>% 
                                rownames_to_column("SAMPID") %>% 
                                select(SAMPID,tissue,natalgroup), by="SAMPID")

kmann_postnatal_filtered_normcount_meta_df <- assayData(kmann_postnatal_filtered)[["normalizedMatrix"]] %>% 
                    as.data.frame() %>%
                    rownames_to_column(., var = "gene_id") %>%
                    pivot_longer(cols = -c("gene_id"), names_to = "SAMPID", values_to = "norm_count") %>%
                    left_join(pData(kmann_postnatal_filtered) %>% 
                                rownames_to_column("SAMPID") %>% 
                                select(SAMPID,tissue,natalgroup), by="SAMPID")

# we can save the normalized counts and sample info separately 
# data is saved in long format which is easier to import and plot in R/python
# let me know if you need in wide format (genes as rows, and samples as columns)

write.table(kmann_prenatal_filtered_normcount_meta_df, file=gzfile("kmann_prenatal_filtered_normcount_tissue_info_brainmerged_v7.tsv.gz"), sep="\t", row.names = F, col.names = T, quote = F)
write.table(kmann_postnatal_filtered_normcount_meta_df, file=gzfile("kmann_postnatal_filtered_normcount_tissue_info_brainmerged_v7.tsv.gz"), sep="\t", row.names = F, col.names = T, quote = F)
```


## TiP4: 

For this version of tissue specific we calculate as follows:

 * Reference group by natal state (i.e. median_all and iqr_all would encompass ALL samples from a specific natal state - as samples are already split based on natal state)
 * TiP calculation for a gene in prenatal-brain would be as below, so that TiP values can be compared between pre- vs post-natal samples
 * ((median of all pre-natal brain) - (median of all pre-natal non-brain samples)/(IQR of all pre-natal samples))

```{r tip, include=T, echo=T, error=F, warning=F, message=F}

# Let us write some functions which parallelizes the metrics calculations
calc_metrics_and_save <- function(df, num_clusters=50, saveFileName="temp"){
      cluster = new_cluster(num_clusters)
      
      #split the data-by gene
      df_gene = df %>% 
      group_by(gene_id) %>%
      partition(cluster)

      # gene-level metrics
      gene_metrics=df_gene %>%
      summarise(
            iqr_all=IQR(norm_count)
      )
      gene_metrics = gene_metrics %>% collect()

      #split the data-by gene and tissue
      df_gene_tissue = df %>% 
      group_by(gene_id, tissue) %>%
      partition(cluster)

      # copy df to another as we need it to calculate median_non_tissue
      tmpdf = df
      cluster_library(cluster,"dplyr")
      cluster_library(cluster,"purrr")
      cluster_copy(cluster,"tmpdf")
      
      gene_tissue_metrics = df_gene_tissue %>%
      summarise(
            median_tissue=median(norm_count),
            median_nontissue=median(tmpdf$norm_count[tmpdf$gene_id == cur_group()$gene_id & tmpdf$tissue != cur_group()$tissue])
      )
      
      gene_tissue_metrics = gene_tissue_metrics %>% collect()

      rm(cluster)
      gc()

      # summarize and calc TiP4
      metrics_df = gene_metrics %>% 
      left_join(gene_tissue_metrics, by=c("gene_id"))

      write.table(metrics_df, file=gzfile(paste0(saveFileName,"_metrics.tsv.gz")),sep="\t", row.names = F, col.names = T, quote = F)
      
      tip_metrics = metrics_df %>%
      group_by(gene_id,tissue) %>%
      summarise(tip4=(median_tissue-median_nontissue)/iqr_all)

      write.table(tip_metrics, file=gzfile(paste0(saveFileName,"_TiP.tsv.gz")),sep="\t", row.names = F, col.names = T, quote = F)
}

calc_metrics_and_save(kmann_prenatal_filtered_normcount_meta_df, num_clusters=100, saveFileName="kmann_prenatal_filtered_brainmerged_v7")
calc_metrics_and_save(kmann_postnatal_filtered_normcount_meta_df, num_clusters=100, saveFileName="kmann_postnatal_filtered_brainmerged_v7")

```
