---
title: "processGTExdata_YARN_TissueSpecificity_v2_All"
output:
  html_document:
  toc: TRUE
  toc_float: TRUE
  keep_md: FALSE
---

```{r setup, include=T, echo=T,error=F,warning=F,message=F}
#######
# Who: Siva
# When: Jan 2022
# What:
# - Load GTEx v8 data (raw; obtained from v1 folder)
# - Filter out the testis, and cell derived samples (Cells - EBV-transformed lymphocytes; Cells - Cultured fibroblasts)
# - Merge the tissue sub-locations for the selected ones (see below for details)
# - Filter low expressed genes
# - apply YARN normalization for the sub-tissue
# - preferential expression calculated using Eqn. 4 and 5 of publication https://doi.org/10.1016/j.celrep.2017.10.001
# - calculate preferential expression 
#######

knitr::opts_chunk$set(echo = F, 
                      cache = F, 
                      warning = F)
options(stringsAsFactors = F)

library(yarn)
library(readr)
library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(reshape2)

```

## Import raw and filter data

We import the raw gtex data from the v1 and filter out the cell derived samples namely Cells - EBV-transformed lymphocytes, Cells - Cultured fibroblasts.

```{r prepData, include=T, echo=T, error=F, warning=F, message=F}
# Just check if pre-processed gtex data exists and load it if available
gtex_raw_no_cells_file_name <- paste0("gtex_raw_no_cells_expSetObj.RData")
if(!file.exists(gtex_raw_no_cells_file_name)){
  load("../v1/gtex_raw_expSetObj.RData")
  # remove the cell derived samples and Testis
  gtex_raw_no_cells <- filterSamples(gtex_raw, ids = c("Cells - EBV-transformed lymphocytes", "Cells - Cultured fibroblasts", "Testis"), groups = "SMTSD", keepOnly =  FALSE)
  
  save(gtex_raw_no_cells, file = gtex_raw_no_cells_file_name)
} else {
  load(gtex_raw_no_cells_file_name)
}

# remove original file and free up mem
rm(gtex_raw)
gc()
```

## Create different groupings of sub-tissue locations

We create the following groups of sub-tissues which were identified based on the v1 analysis (refer to Tissue sub-locations section of the v1 report). The tissues which are not mentioned below did not have sub tissue locations or we did not identify useful clusters in the v1 analysis.  

 * Adipose tissue (2: subcutaneous, visceral); 
 * Blood vessel (3: Aorta, coronary, tibial); 
 * Brain (3: Cerebellum, Basal ganglia, Others);
 * Colon (2: sigmoid, transverse); 
 * Esophagus (2: mucosa, muscularis + gastroesophagal junction); 
 * Heart (2: left ventricle; atrial appendage); 
 * Kidney (2: cortex; medulla); 
 * Skin (1: Not sun exposed + sun exposed);

For the ease of processing and trackability let us create one additional column named SMTSD_New (for the grouping)  in the phenoData slot of the expression set object; SMTSD column already has the info of all the sub-sampling sites.

```{r create_subgroupings, include=T, echo=T, error=F, warning=F, message=F}
gtex_raw_no_cells$SMTSD_New <- gtex_raw_no_cells$SMTSD

# adipose tissue, blood vessel, colon, and heart already have the groupings as we intend to have.

# We merge the cerebellum and cerebellar hemisphere samples
# We merge the caudate, accumbens, and putamen of basal ganglia as one group
# we merge the rest of the brain sub-tissues as others
pData(gtex_raw_no_cells)[grep("brain",pData(gtex_raw_no_cells)$SMTSD, ignore.case = TRUE, fixed = FALSE),]$SMTSD_New <- "Brain - Others"
pData(gtex_raw_no_cells)[grep("cerebell",pData(gtex_raw_no_cells)$SMTSD, ignore.case = TRUE, fixed = FALSE),]$SMTSD_New <- "Brain - Cerebellum"
pData(gtex_raw_no_cells)[grep("basal ganglia",pData(gtex_raw_no_cells)$SMTSD, ignore.case = TRUE, fixed = FALSE),]$SMTSD_New <- "Brain - Basal ganglia"

# we merge cervix as one tissue
pData(gtex_raw_no_cells)[grep("cervix",pData(gtex_raw_no_cells)$SMTSD_New, ignore.case = TRUE, fixed = FALSE),]$SMTSD_New <- "Cervix Uteri"

# we merge in Esophagus muscularis and gastroesophagal junction as gastroesophageal junction
pData(gtex_raw_no_cells)[grep("muscularis",pData(gtex_raw_no_cells)$SMTSD_New, ignore.case = TRUE, fixed = FALSE),]$SMTSD_New <- "Esophagus - Gastroesophageal Junction"

# we merge kidney as one tissue
pData(gtex_raw_no_cells)[grep("kidney",pData(gtex_raw_no_cells)$SMTSD_New, ignore.case = TRUE, fixed = FALSE),]$SMTSD_New <- "Kidney"

# we merge skin as one tissue
pData(gtex_raw_no_cells)[grep("skin",pData(gtex_raw_no_cells)$SMTSD_New, ignore.case = TRUE, fixed = FALSE),]$SMTSD_New <- "Skin"

```

## Filtering low expressed genes

We filter out genes from subsequent analysis which are low expressed (at least 1 read in half of the samples per tissue).


```{r filter_genes_S3, include=T, echo=T, error=F, warning=F, message=F}
gtex_filtered <- filterLowGenes(gtex_raw_no_cells, "SMTSD_New")

```

The number of genes in our dataset before and after filtering are `r dim(gtex_raw_no_cells)[1]` and `r dim(gtex_filtered)[1]`. 

## Normalization

We can perform the tissue aware normalization. Using the plots shown below, we can observe the effect of normalization on all the samples. 

```{r normalize_genes_S3, include=T, echo=T, error=F, warning=F, message=F}
plotDensity(gtex_filtered, "SMTSD_New", main=expression('log'[2]*' raw expression'),
             xlab="Read count", ylab="Density")
gtex_filtered <- normalizeTissueAware(gtex_filtered,"SMTSD_New")
plotDensity(gtex_filtered, "SMTSD_New", normalized=TRUE, main="Normalized - All tissue subgroups", xlab="Read count", ylab="Density")
```

## Calculating specificity score

We extract here the required normalized count matrix and the associated sample metadata (a.k.a phenoData), specifically sample id and the tissue name.

```{r calc_specificity_S3, include=T, echo=T, error=F, warning=F, message=F}

# get required columns and join them from assayData and phenoData of the ExpressionSet object
gtex_filtered_normcount_meta_df <- assayData(gtex_filtered)[["normalizedMatrix"]] %>% 
					as.data.frame() %>%
					rownames_to_column(., var = "gene_id") %>%
					pivot_longer(cols = -c("gene_id"), names_to = "SAMPID", values_to = "norm_count") %>%
					left_join(pData(gtex_filtered) %>% 
							select(SAMPID,SMTSD_New), by="SAMPID")

# we can save the normalized counts and sample info separately 
# data is saved in long format which is easier to import and plot in R/python
# let me know if you need in wide format (genes as rows, and samples as columns)

# commenting out below statements as it produces a rather huge file and this may not be needed

# write.table(gtex_filtered_normcount_meta_df, file="gtex_all_tissue_subgroups_filtered_normcount_tissue_info.tsv", 
#	    sep="\t", row.names = F, col.names = T, quote = F)

# In this step, we calculate the tissue specific medians; all tissues medians; 
# inter-quartile range of all tissues; and the calculate the specificity score
gtex_filtered_norm_specificity_info <- gtex_filtered_normcount_meta_df %>%
					group_by(gene_id) %>%
					summarise(median_all=median(norm_count), iqr_all=IQR(norm_count)) %>%
					left_join(gtex_filtered_normcount_meta_df %>%
						  group_by(gene_id,SMTSD_New) %>%
						  summarise(median_tissue=median(norm_count)), by="gene_id") %>%
					mutate(specificity_score=round((median_tissue-median_all)/iqr_all,2))


# we can now save these data as tables

write.table(gtex_filtered_norm_specificity_info, file="gtex_all_tissue_subgroups_filtered_norm_specificity_info.tsv", 
	    sep="\t", row.names = F, col.names = T, quote = F)
```

## Distribution of specificity score

Let us now take a look at the distribution of the specificity score. 

```{r distro_specificity_S3, include=T, echo=T, error=F, warning=F, message=F}

# basic summary of specificity scores; if there is no infinity found, we can be happy that we did not end up with a zero IQR value

print(summary(gtex_filtered_norm_specificity_info$specificity_score))

ggplot(gtex_filtered_norm_specificity_info, aes(specificity_score))+
	geom_histogram()+
	theme_bw()+
	labs(title="Specificity score distribution - All tissue subgroups", x="Specificity score", y="Count")

```

Next, let us have a look at the number of genes specific in at least one tissue for different specificity cut-offs (1 to 3 with a step size of 0.25). 


```{r distro_specificity_counts_S3, include=T, echo=T, error=F, warning=F, message=F}

n_genes_specificity_score_cuts_tissue_subgroups <- data.frame(sp_cut_off=c(seq(1,3,0.25)), 
					  tissue_subgroups=sapply(c(seq(1,3,0.25)), function(x) { 
								  	  gtex_filtered_norm_specificity_info %>% 
									  	filter(specificity_score > x) %>%
									        select(gene_id) %>%	
									  	n_distinct() }
					  )
					  )

print(n_genes_specificity_score_cuts_tissue_subgroups)

n_genes_specificity_score_cuts_tissue_subgroups %>% 
	pivot_longer(-c(sp_cut_off),names_to = "type", values_to = "count") %>% 
	ggplot(aes(x=sp_cut_off, y=count)) + 
		geom_bar(stat="identity", position="dodge") + 
		theme_bw()+
		labs(x="Specificity score cut-offs", y="# of distinct genes higher than the cut-off")

```

## Multiplicity of the tissue specific genes

We can get in how many tissues each gene is found to be specific. A boxplot (and a histogram) of the distribution of the mulitiplicity values is also shown, just to get an idea of how often are genes found to specific for multiple tissues. For these calculations, I have used a specificity cut-off of 2. 

```{r multiplicity_counts_S3, include=T, echo=T, error=F, warning=F, message=F}

gene_multiplicity_tissue_subgroups <- gtex_filtered_norm_specificity_info %>% filter(specificity_score > 2) %>% count(gene_id)

write.table(gene_multiplicity_tissue_subgroups, file="gtex_all_tissue_subgroups_specificity_2_filtered_gene_multiplicty.tsv", 
	    sep="\t", row.names = F, col.names = T, quote = F)

gene_multiplicity_tissue_subgroups %>%
	ggplot(aes("", n)) +
	geom_boxplot() +
	theme_bw() + 
	labs(x="Distribution", y="Multiplicity", title="Multiplicity distribution - All tissue subgroups")

gene_multiplicity_tissue_subgroups %>%
	ggplot(aes(n)) +
	geom_histogram() +
	theme_bw() +
	labs(x="Multiplicity", y="Number of genes with such multiplicity", title="All tissue subgroups" )

```

## Saving the filtered data

```{r save_all_S3, include=T, echo=T, error=F, warning=F, message=F}
save(gtex_filtered, file="gtex_all_tissue_subgroups_filtered_normalized_expsetobj.RData")
```
